Actuator line modeling of vertical-axis turbines

### 2016-05-13 paper.tex
Add concluding sentence
Move inflow velocity detection up and slim down theory

The mean near-wake structure of vertical-axis turbine has been shown to be
largely dominated by the effects of tip vortex shedding, which induces levels of
recovery due to vertical advection significantly larger than those from
turbulent fluctuations~\cite{Bachant2015-JoT}---an effect not seen in HAT wakes.
As interest shifts to designing and analyzing arrays of VATs, it is necessary to
determine the effectiveness of various numerical modeling techniques to
replicate VAT near-wake dynamics, such that wake recovery is accurately
predicted, leading to accurate assessment of optimal array spacing.

@@

The actuator line model is based on the classical blade element theory combined
with a Navier--Stokes description of the flow field. The ALM treats turbine
blades as lines of blade elements, for which 2-D profile lift and drag
coefficients are given. For each blade element, relative flow velocity and angle
of attack are computed by adding the vectors of relative blade motion and the
local fluid velocity, a diagram of which is shown in Figure~\ref{fig:vectors}.

@@

The kinematics of a CFT are parameterized by the tip speed ratio $\lambda =
\omega R/U_\infty$, where $\omega$ is the shaft's angular velocity, $R$ is the
maximum rotor radius, and $U_\infty$ is the free stream velocity. Using only
geometric considerations---that is, ignoring the effect of the rotor forcing and
assuming the inflow velocity to be equal to that of the free stream---the
geometric angle of attack $\alpha$ and relative velocity magnitude
$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor
azimuthal angle $\theta$, which is plotted for various tip speed ratios in
Figure~\ref{fig:geom-alpha-urel}.

    \includegraphics[width=0.75\textwidth]{CFT-vectors_alpha_deg_urel_geom}

    \caption{Geometric angle of attack (left) and relative velocity (right)
        versus azimuthal angle at various tip speed ratios.}

    \label{fig:geom-alpha-urel}
\end{figure}

Inspecting Figure~\ref{fig:geom-alpha-urel}, it is apparent that CFTs are
inherently unsteady machines, encountering large amplitude and rapid
oscillations in angle of attack, often exceeding the rotor blades' static stall
angles in typical operation. The unsteadiness can be characterized by a reduced
frequency~\cite{Leishman2006}
\begin{equation}
    k = \frac{\omega c}{2 U_\infty},
\end{equation}
which assumes the free stream velocity is constant. Unsteady effects begin to
become significant for $k > 0.05$, and can become dominant for $k \ge 0.2$.

@@

\section{Determining inflow velocity}

@@

The ALM provides a more physical flow description compared to momentum and
potential flow vortex models, at a reasonable cost. The ALM also drastically
reduces computational effort compared to blade-resolved CFD, while maintaining
the unsteadiness of the wake not resolved by a conventional actuator disk. When
combined with RANS the ALM will allow VAT array simulations on individual PCs,
and with high performance computing and LES the ALM is one of the highest
fidelity array modeling tools available.

>>>>

The mean near-wake structure of vertical-axis turbine has been shown to be
largely dominated by the effects of tip vortex shedding, which induces levels of
recovery due to vertical advection significantly larger than those from
turbulent fluctuations~\cite{Bachant2015-JoT}---an effect not seen in axial-flow
or horizontal-axis turbine (AFT or HAT) wakes. As interest shifts to designing
and analyzing arrays of VATs, it is necessary to determine the effectiveness of
various numerical modeling techniques to replicate VAT near-wake dynamics, such
that wake recovery is accurately predicted, leading to accurate assessment of
optimal array spacing.

@@

The actuator line model is based on the classical blade element theory combined
with a Navier--Stokes description of the flow field. The ALM treats turbine
blades as lines of blade elements, for which 2-D profile lift and drag
coefficients are given. For each blade element, relative flow velocity
$\vec{U}_\mathrm{rel}$ and angle of attack $\alpha$ are computed by adding the
vectors of relative blade motion $-\omega r$, where $\omega$ is the rotor
angular velocity and $r$ is the blade element radius, and the local inflow
velocity $\vec{U}_\mathrm{in}$, a diagram of which is shown in
Figure~\ref{fig:vectors}.

@@

The kinematics of a CFT are parameterized by the tip speed ratio $\lambda =
\omega R/U_\infty$, where $R$ is the maximum rotor radius, and $U_\infty$ is the
free stream velocity. In contrast to an AFT, which under ideal conditions can be
considered a steady machine, CFT blades encounter large amplitude and rapid
oscillations in angle of attack, often exceeding the rotor blades' static stall
angles in typical operation, which makes their blade loading and performance
more difficult to predict. The unsteadiness can be characterized by a reduced
frequency~\cite{Leishman2006}
\begin{equation}
    k = \frac{\omega c}{2 U_\infty},
\end{equation}
which assumes the free stream velocity is constant. Unsteady effects begin to
become significant for $k > 0.05$, and can become dominant for $k \ge 0.2$.

@@

@@

The ALM provides a more physical flow description compared to momentum and
potential flow vortex models, at a reasonable cost. The ALM also drastically
reduces computational effort compared to blade-resolved CFD, while maintaining
the unsteadiness of the wake not resolved by a conventional actuator disk. When
combined with RANS the ALM will allow VAT array simulations on individual PCs,
and with high performance computing and LES the ALM is one of the highest
fidelity array modeling tools available. Ultimately, the ALM will help reduce
dependence on expensive physical modeling to optimize VAT array layouts.
